---
title: "tidymodels"
author: "Ravi Brenner"
date: "2025-06-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

Following along with the tidymodels tutorial found here: https://www.tidymodels.org/start/models/

I have a lot of R experience, and some experience with caret and tidymodels too, so hopefully this will be easy for me.

# Build a model
```{r}
library(tidymodels)
library(readr)       # for importing data
library(broom.mixed) # for converting bayesian models to tidy tibbles
library(dotwhisker)  # for visualizing regression results

theme_set(theme_bw())
```

Read in sea urchins data
```{r}
urchins <- read_csv("https://tidymodels.org/start/models/urchins.csv") |>
  setNames(c("food_regime","initial_volume","width")) |>
  mutate(food_regime = factor(food_regime, levels = c("Initial", "Low", "High")))

head(urchins)
str(urchins)
```

We can quickly plot this data, since it isn't very large

```{r}
ggplot(urchins,
       aes(x = initial_volume, y = width , color = food_regime)) +
  geom_point() + 
  geom_smooth(method = lm, se = FALSE) + 
  scale_color_viridis_d(option = "plasma", end = 0.7) 
```

So there's a generally positive relationship, and it appears to be impacted by food_regime

## fitting a simple model

Creating the model in tidymodels means creating a pipeline of components

```{r}
lm_mod <- linear_reg() 
```

```{r}
lm_fit <- lm_mod |>
  fit(width ~ initial_volume * food_regime, data = urchins)

tidy(lm_fit)
```

```{r}
tidy(lm_fit) |>
  dotwhisker::dwplot(dot_args = list(size = 2,color = "black"),
                     whisker_args = list(color = "black"),
                     vline = geom_vline(xintercept = 0, color = "gray",linetype = 2))
```

So we very easily fit a model. `fit` from parsnip is clearly very similar to `train` from caret

## Predicting with the model

```{r}
new_points <- expand_grid(initial_volume = 20, 
                          food_regime = c("Initial", "Low", "High"))

mean_pred <- predict(lm_fit, new_data = new_points)
mean_pred

conf_int_pred <- predict(lm_fit, new_data = new_points,
                         type = "conf_int")
conf_int_pred

plot_data <- new_points |>
  bind_cols(mean_pred) |>
  bind_cols(conf_int_pred)
```

```{r}
ggplot(plot_data, aes(x = food_regime)) + 
  geom_point(aes(y = .pred)) + 
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = .2) + 
  labs(y = "urchin size")
```

Very simple, and familiar from working with caret.

## use a different engine

We can train more complicated models using other engines. here is a stan_glm model

```{r}
prior_dist <- rstanarm::student_t(df = 1)

set.seed(123)

# creating the model itself
bayes_mod <- linear_reg() |>
  set_engine("stan",
             prior_intercept = prior_dist,
             prior = prior_dist)

bayes_fit <- bayes_mod |>
  fit(width ~ initial_volume * food_regime, data = urchins)
```

```{r}
tidy(bayes_fit, conf.int = T)
```

We can now recycle our code from earlier

```{r}
bayes_plot_data <- new_points |>
  bind_cols(predict(bayes_fit, new_data = new_points)) |>
  bind_cols(predict(bayes_fit, new_data = new_points, type = "conf_int"))
```

```{r}
ggplot(bayes_plot_data, aes(x = food_regime)) + 
  geom_point(aes(y = .pred)) + 
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = .2) + 
  labs(y = "urchin size") + 
  ggtitle("Bayesian model with t(1) prior dist")
```

# Preprocess data with recipes

# Evaluate model with resamples

# Tune model parameters

# Case study

# My personal case study